The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-11-15 script/en/chapters/paper_and_pencil.tex
CT script: Minor corrections, mostly layout, de + en; final now for CT 1.4.30 Beta04
examples on the shift and Caesar ciphers using Sage

@@

\bibitem[Schmeh2007]{pp:Schmeh2007}  \index{Schmeh 2007}
        Klaus Schmeh, \\
        {\em Codeknacker gegen Codemacher. Die faszinierende Geschichte der Verschl\"usselung.},\\ 
        W3L Verlag Bochum, 2. Auflage 2007.\\
	This is most current among the books dealing in a comprehensive manner
	with the history of cryptology. It contains a small collection of solved
	and unsolved crypto riddles. One of the challenges deals with a double
	column transposition using two long keys, which are differently in addition.

>>>>

% ---------------------------------------------------------------------------
\newpage
\subsection{Shift cipher}

The shift cipher can be thought of as a generalization of the Caesar
cipher. While the Caesar cipher restricts us to shift exactly three
positions along an alphabet, the shift cipher allows us to shift any
number of positions along the alphabet. The shift cipher is
implemented in the Sage class
\begin{center}
\verb!sage.crypto.classical.ShiftCryptosystem!
\end{center}
In the following example, we construct a shift cipher over the capital
letters of the English alphabet. We then encrypt a plaintext by
shifting it 12 positions along the alphabet. Finally, we decrypt the
ciphertext and make sure that it is indeed the original plaintext.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: shiftcipher = ShiftCryptosystem(AlphabeticStrings()); shiftcipher
Shift cryptosystem on Free alphabetic string monoid on A-Z
sage: plaintext = shiftcipher.encoding("Shift me any number of positions.")
sage: plaintext
SHIFTMEANYNUMBEROFPOSITIONS
sage: key = 12  # shift by 12 positions
sage: 
sage: # shift the plaintext by 12 positions to get the ciphertext
sage: ciphertext = shiftcipher.enciphering(key, plaintext)
sage: ciphertext
ETURFYQMZKZGYNQDARBAEUFUAZE
sage: 
sage: # decrypt the ciphertext and ensure that it is the original plaintext
sage: P = shiftcipher.deciphering(key, ciphertext); P
SHIFTMEANYNUMBEROFPOSITIONS
sage: P == plaintext
True
\end{Verbatim}
\caption{A shift cipher over the upper-case letters of the English alphabet.}
\end{sagecode}

The Caesar cipher is simply a shift cipher whose shifting key is 3. In
the next example, we use the shift cipher to create a Caesar cipher
over the capital letters of the English alphabet. We then encrypt a
plaintext using the Caesar cipher, and then proceed to decrypt the
resulting ciphertext. Finally, we ensure that the decrypted result is
the same as our original plaintext.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # create a Caesar cipher
sage: caesarcipher = ShiftCryptosystem(AlphabeticStrings())
sage: plaintext = caesarcipher.encoding("Shift the alphabet by three positions to the right.")
sage: plaintext
SHIFTTHEALPHABETBYTHREEPOSITIONSTOTHERIGHT
sage: 
sage: # shift the plaintext by 3 positions
sage: key = 3
sage: ciphertext = caesarcipher.enciphering(key, plaintext)
sage: ciphertext
VKLIWWKHDOSKDEHWEBWKUHHSRVLWLRQVWRWKHULJKW
sage: 
sage: # decrypt the ciphertext and ensure that it is the original plaintext
sage: P = caesarcipher.deciphering(key, ciphertext)
sage: P
SHIFTTHEALPHABETBYTHREEPOSITIONSTOTHERIGHT
sage: P == plaintext
True
\end{Verbatim}
\caption{Constructing the Caesar cipher using the shift cipher.}
\end{sagecode}

@@

\bibitem[Schmeh2007]{pp:Schmeh2007}  \index{Schmeh 2007}
        Klaus Schmeh, \\
        {\em Codeknacker gegen Codemacher. Die faszinierende Geschichte der Verschl\"usselung},\\ 
        W3L Verlag Bochum, 2. Auflage 2007.\\
	This is most current among the books dealing in a comprehensive manner
	with the history of cryptology. It contains a small collection of solved
	and unsolved crypto riddles. One of the challenges deals with a double
	column transposition using two long keys, which are differently in addition.
